---
title: "sp2eco - Getting started"
vignette: >
  %\VignetteIndexEntry{Getting Started with sp2eco}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format: html
---

```{r}
#| label: setup
library(sp2eco)
```

## 1. Introduction
<p align="justify">
The folder in which this notebook was included contains the data and code that 
are required to execute this example. Notably, the source code for **GlobES**, 
and R package containing the functions required to execute the workflow. In 
addition, the following folder structure is given:
<li><b>input</b>. Provides example input data required to run this example.</li> 
<li><b>output</b>. Intended to preserve generated outputs.</li>
<li><b>info</b>. Includes files with key information guiding our analysis, such as 
species names and their range map identifiers, species-habitat associations and 
elevation ranges, and a listing of all ecosystem types, and their thematic groups
and land cover classes.</li>
As an example, we will here apply our workflow to derive ecosystem occurrences for 
Mediterranean scrub using species observations from the Global Biodiversity Information 
Facility (GBIF). The focal ecosystem type is described in the habitat typology of the 
IUCN Red List of Species, which serves as basis for globally-applicable species threat 
assessments (<a href="https://www.iucnredlist.org/resources/habitat-classification-scheme">see here for details</a>). 
The code below defines required arguments, installs package dependencies, and loads needed information files.
</p>

```{r}
# arguments ----
#-----------------------------------------------------------------------------#
# target resolution
map_res = 0.00833

# class identifier of Mediterranean Shrubland
cid = "0308"

# estimate decay rate value for 2-10 habitat associations
# NOTE: 2-10 is maximum number at which a weight of 0.01 is given
lambda = sapply(2:10, function(d) 1/(d-1)*log(0.01/1), USE.NAMES=F)

# Niche Rover parameters (used in sensitivity analysis)
nsamples = 1000 # number of MCMC draws
nreps = 10 # number of metrics to derive
niter = 100 # number of sub-sample iterations

# install/load data packages ----
#-----------------------------------------------------------------------------#

# list of packages required (except GlobES)
pkgs = c("readr","nicheROVER","clusterSim","terra","plyr")
for (p in pkgs) install.packages(p)

install.packages("GlobES_0.1.1.tar.gz", method="source")

# load dependencies
for (p in c("GlobES",pkgs)) require(p, character.only=T, quietly=T)

# load information files ----
#-----------------------------------------------------------------------------#

# information file on the IUCN typology
typology = read_csv(file.path("info", "iucn_typology.csv"), show_col_types=F)

# species habitat associations
habitat = read_csv(file.path("info", "iucn_habitat.csv"), show_col_types=F)

# target species list
target = read_csv(file.path("info", "iucn_species.csv"), show_col_types=F)

# elevation ranges
elev_ranges = read_csv(file.path("info", "iucn_elevation.csv"), show_col_types=F)
```

## 2. Methods
<p align="justify">
This demonstration is organized in two major groups, as per the description 
of our framework in the main manuscript. First, we apply tests controlling 
for errors in reported coordinate locations (2.1, same as in the manuscript). 
Second, we estimate ecosystem occurrences probabilities regionally, and use 
those estimates to compute parameters for a Dirichlet conjugate prior distribution. 
(section 2.2). These parameters are then used to update local likelihoods of a 
derived ecosystem occurrence being correctly labeled. Furthermore, entropy-based 
thresholds are inferred from the probability maps to filter ecosystem occurrence 
probabilities (2.3). To demonstrate this workflow, the following data are provided:
</p>

```{r, eval=F}
# candidate GBIF observations for Mediterranean shrubland
obs = read_csv(file.path("input", "spObs", "spObs-gbif.csv"), show_col_types=F)

# Digital Elevation Model (DEM)
elevation = rast(file.path("input","dem.tif"))

# land cover map
landcover = rast(file.path("input", "GLAD1k", "landCover-short_1km.tif"))
```

<p align="justify>
Note that the provided data include range maps that already rasterized, and a 
spatial resolution of 2-degrees. The originals cannot be shared because their 
user agreement prevents redistribution. However, these can be directly downloaded 
from the website of the <a href="https://www.iucnredlist.org/">IUCN Red List of Species</a>.
</p>
<p align="justify">
Due to the large data volume required to reproduce our showcase examples, the 
provided range map data account only for species inhabiting forest ecosystems. 
To then demonstrate how species observations are derived, we focus on Mediterranean 
shrubland and a spatial resolution of 1-km resolution. However, only sections 
2.1 and 2.2 (the key section demonstrating our workflow) is fully executable. 
This is because section 2.3 (i.e., sensitivity analysis) requires more data to 
fulfill all filtering scenarios, as well as the processing of additional ecosystem types.
</p>

## 2.1. Location filters

```{r}
# setup analysis ----
#-----------------------------------------------------------------------------#

# focal species
species = unique(obs$species)

# obtain range identifiers for each species
rid = target$range_id[match(species,target$scientificname)]

# construct path to range maps of focal species
g_ranges = file.path("input", "rangeMap_mask", paste0(rid, ".tif"))

# reduce focus to species with range maps
ind = which(file.exists(g_ranges))
g_ranges = rast(g_ranges[ind])
species = species[ind]

# name file path vector
names(g_ranges) = species

# isolate elevation ranges of each species
e_ranges = elev_ranges[match(species,elev_ranges$scientificName),]
e_ranges = lapply(1:nrow(e_ranges), function(i) unlist(e_ranges[i,2:3]))
names(e_ranges) = species

# apply tests ----
#-----------------------------------------------------------------------------#

flag = location_filter(
  
  # observation coordinates
  cbind(obs$decimallongitude,obs$decimallatitude),
  
  # coordinate uncertainty
  obs$coordinateuncertaintyinmeters, 
  
  # names of the observed species
  obs$species, 
  
  # iso-2 code of country where observation was made
  obs$countrycode,
  
  # target pixel resolution
  map_res,
  
  # named paths to range map masks of each unique species  
  g_ranges, 
  
  # elevation ranges of each observed species
  e_ranges,
  
  # elevation raster
  elevation, 
  
  # assumes that g_ranges is a stack of range maps
  method="range"

)

# add metadata to observations
obs$cell = flag$cell # cell in global grid with res. of 0.00833-degrees
obs$coordinateuncertaintyinmeters = flag$precision ## updated precision
obs$resolution = flag$resolution # resolution linked to 'cell' (not needed here)

# derive candidate ecosystem occurrences ----
#-----------------------------------------------------------------------------#

# exclude invalid observations
obs = obs[flag$flag,]

# isolate potential Mediterranean Shrubland occurrences
sp = unique(habitat$scientificName[which(habitat$class_id == cid)])
obs = obs[which(obs$species %in% sp),]

# infer level of habitat specialization for each species
nh = ddply(habitat[habitat$scientificName %in% obs$species,], .(scientificName), 
           summarize, n=length(unique(class_id)))

# assign habitat count to each species observation
obs$n_habitats = nh$n[match(obs$species,nh$scientificName)]

# reduce observations to unique cell-year combinations
obs = ddply(obs, .(cell,year), summarize, 
            
            # mean coordinates
            x=mean(decimallongitude), y=mean(decimallatitude), 
            
            # mean precision
            coordinateuncertainty=mean(coordinateuncertaintyinmeters),
            
            # name of the species from which the observations was obtained
            species=species[which.min(n_habitats)], 
            
            # number of species observation
            n_species=length(unique(species)),
            
            # highest degree of habitat specialization
            n_habitats=min(n_habitats))

# test land cover condition ----
#-----------------------------------------------------------------------------#

e = extract(landcover, obs[,c("x","y")], ID=F)[,1]
obs$lc_check = e > 0
```

## 2.2. Controlling for thematic uncertainties

First, we map ecosystem occurrence probabilities at a coarse spatial scale.
```{r, eval=F}
# prepare arguments ----
#-----------------------------------------------------------------------------#

# list range maps encoded with the number of habitat associations
files = list.files(file.path("input", "rangeMap_habitat"), ".tif", full.names=T)

# extract class code from file names
eco = do.call(rbind, strsplit(basename(files), "_"))[,1]

# derive per-pixel Dirichlet distribution parameters (alpha) ----
#-----------------------------------------------------------------------------#

# map probabilities for each lambda 
p_path = file.path("output", "occ_probabilities")
dir.create(p_path)
alpha = range_prior(files, eco, lambda, p_path, verbose=F)

# compute total (used to re-normalized updated values)
total_alpha = app(alpha, sum)

# update local votes ----
#-----------------------------------------------------------------------------#

# derive weighted votes
w_votes = weighted_sum(obs$n_habitats, lambda)

alpha = extract(alpha[[which(names(alpha) == cid)]], obs[,c("x","y")], ID=F)[,1]
total_alpha = extract(total_alpha, obs[,c("x","y")], ID=F)[,1]

# lower confidence interval
obs$lower_ci = ((w_votes$mean-w_votes$var) + alpha) /
  ((w_votes$mean-w_votes$var) + total_alpha)

# mean
obs$p_mean = (w_votes$mean + alpha) / (w_votes$mean + total_alpha)

# higher confidence interval
obs$upper_ci = ((w_votes$mean+w_votes$var) + alpha) /
  ((w_votes$mean+w_votes$var) + total_alpha)

# record number of specialists per cell ----
#-----------------------------------------------------------------------------#

# list range maps encoded with the number of habitat associations
files = list.files(file.path("input", "rangeMap_habitat"), 
                   "0308_.*.tif", full.names=T)

# sum specialist ranges
r = app(rast(files), function(x) sum(x == 1))

# test if number of regional specialist occurrences is > 1
obs$specialist_check = extract(r, obs[,c("x","y")], ID=F)[,1] > 1
```

<p align="justify">
Second, for each mean probability map, we estimate entropy-based thresholds to 
filter low-probability areas. With this threshold, we can identify a final set 
of occurrence records for Mediterranean scrub. This is obtained by imposing the 
various quality criteria tested in our framework The preserved species observations 
must: i) overlap supporting land cover, ii) match or exceed the regional probability 
threshold, and iii) occur in a region where at least two habitat specialists 
within a 2-degree cell.
</p>
</p>

```{r, eval=F}
# path to output raster
p_map = rast(file.path(p_path, "0308_probability_mean.tif"))

# estimate threshold 
e_threshold = entropy_threshold(p_map)

# detect observations matching condition
obs$p_check = FALSE
obs$p_check[which(extract(p_map, obs[,c("x","y")], ID=F)[,1] > e_threshold)] = TRUE

# save candidates (to apply sensitivity analysis)
write_csv(obs, file.path("output", "0308-candidate_occurrences.csv"))

# conditions to be applied to isolate valid observations
obs = obs[which(obs$lc_check & obs$p_check & obs$specialist_check),]

# save final set of observations
write_csv(obs, file.path("output", "0308-processed_occurrences.csv"))
```

## 2.3. Sensitivity analysis

<p align="justify">
To evaluate the performance of our framework, we conducted two sets of tests. 
First, we computed the probability of encountering derived ecosystem occurrences 
within the bioclimatic niche of reference data on the target ecosystems. Second, 
we evaluated the separability of ecosystem occurrence records fo different types. 
We evaluated these metrics for the following filtering options:
<ul>
<li><b>N - </b>No filtering</li>
<li><b>S - </b>Number of specialists within a grid cell.</li>
<li><b>L - </b>Land cover support.</li>
<li><b>T - </b>Entropy-based thematic uncertainty filter</li>
<li><b>So - </b>Observations derived from habitat specialists only.</li>
</ul>
We additionally considered combinations of these filters (i.e., S+L, S+T, L+T, S+L+T, So+S+L+T).
</p>
<p align="justify">
To test for bioclimatic niche overlaps, we obtained reference data from the 
sPlot-Open database. The code below shows how reference data were obtained. 
Note that the code generates reference data for all ecosystem types evaluated 
in this study. However, in subsequent demonstrations, we will again focus on 
Mediterranean scrhub, leveraging the output of the previous analyses. The 
code below was used to generate reference data, but samples are already 
in the **input** folder.
</p>

```{r, eval=F}
# download source data of splot
url = "https://idata.idiv.de/MMM/ShowMultimediaData/getFile?path=Datasets%255c3474%255csPlotOpen_header%283%29.txt"
iname = file.path("input", "sPlot", "sPlot-source.txt")
download.file(url, iname, mode="wb")

# load splot data
splot = read_delim(iname, show_col_types=F)

# filters to apply to the splot data for each class
targets = list(
  
  bf=list(id="0101", biome="Boreal zone",cover="Forest"),
  
  tf=list(id="0104", biome="Temperate midlatitudes",cover="Forest"),
  
  df=list(id="0105", biome="Tropics with summer rain",cover="Forest"),
  
  ms=list(id="0308", biome="Subtropics with winter rain",cover="Shrubland"),
  
  tg=list(id="0404",biome="Temperate midlatitudes",cover="Grassland"))

# derive samples
for (target in targets) {
  i = which((splot$Biome %in% target$biome) & (splot[[target$cover]]))
  tmp = splot[i,c("Longitude","Latitude")]
  colnames(tmp) = c("x","y")
  oname = file.path("input", "sPlot", paste0("sPlot-", target$id, ".csv"))
  write.csv(tmp, oname, row.names=F)
}

```

<p align="justify">
We additionally used bioclimatic variables from the CHELSA dataset. These 
data, which consist of 19 bioclimatic variables, were reduced to the 5-leading 
components based on a Principal Components Analysis (PCA) to control for 
multi-colinearity. The leading components accounted for >= 95% fo the variance. 
The code for generating the PCA model is shown below. The code below is used 
for demonstration purposes. An the **output** folder a reduced bioclim dataset 
for the region occupied by the species occurrences used in this example is 
provided, together with a PCA model generated from the original global data. 
</p>

```{r, eval=F}
iname = file.path("output", "bioclim_pca.tif")

if (file.exists(iname)) bioclim = rast(iname) else {
  
  # load bioclimatic data
  files = list.files(file.path("input", "CHELSA"), ".tif", full.names=T)
  
  # load (or download) bioclim variables
  bioclim = tryCatch(rast(files), error=function(e) return(NULL))
  if (is.null(bioclim)) {
    
    files = read.csv(file.path("input", "CHELSA", "CHELSA_source.csv"), header=F)[,1]
    
    pb = txtProgressBar(min=1, max=length(files), style=3)
    for (i in 1:length(files)) {
      o = file.path("input", "CHELSA", basename(files[i]))
      download.file(files[i], o, mode="wb")
      setTxtProgressBar(pb, i)
    }
    
    bioclim = rast(files)
    
  }
  
  # derive 10,000 random samples
  xy = spatSample(bioclim, 10000)

  # apply and save PCA model
  pca_model = prcomp(xy, scale.=T)
  saveRDS(pca_model, file=file.path("ouput", "pca_model.rds"))
  
  # number of components to use in subsequent analyses (>=95%)
  nc = which.min(abs(0.95-summary(pca_model)$importance[3,]))[[1]]
  
  # we can now apply the PCA model to the bioclim data
  bioclim = predict(pca_model, bioclim)
  writeRaster(bioclim, file.path("output", "bioclim_pca.tif"))
  
}
```

<p align="justify">
Below, the code for generating the sensitivity analysis samples. Note that we 
reduce our focus to unique combinations on pixels-observations with respect to 
the resolution of the bioclimatic dataset (i.e., 1-km). The goal is to prevent 
biases due to pseudo-replication. While the bioclimatic data does not change in 
time, and has a fixed resolution, the derive ecosystem occurrences may cluster 
in single pixels (when derived at finer resolutions than 1-km), those sample 
pixels may be sampled in multiple years.
</p>

```{r, eval=F}
# base path for the sensitivity analyses
o_path = file.path("output", "sensitivity")
dir.create(o_path, showWarnings=F)

# names of scenarios (to label samples)
groups = c("N", "S", "L", "T", "S+L", "S+T", "L+T", "S+L+T", "So", "So+S+L+T")

# obtain required data ----
#-----------------------------------------------------------------------------#

# read reference samples
reference_xy = read.csv(file.path("input", "sPlot", "sPlot-0308.csv"))
reference_xy$cell = cellFromXY(bioclim, reference_xy[,c("x","y")])
reference_xy = reference_xy[!duplicated(reference_xy$cell),]

# read target samples
target_xy = read.csv(file.path("output", "0308-candidate_occurrences.csv"))
target_xy = target_xy[complete.cases(target_xy),]
target_xy$cell = cellFromXY(bioclim, target_xy[,c("x","y")])

## summarize observations for each filtering scenario ----
##==========================================================================#

ind = list(

  # no filtering
  1:nrow(target_xy),

  # habitat specialization filter (S)
  which(target_xy$specialist_check),

  # land cover filter (L)
  which(target_xy$lc_check),

  # regional probability filter (T)
  which(target_xy$p_check),

  # land cover + specialization filter ( L + S)
  which((target_xy$lc_check) & target_xy$specialist_check),

  # specialization + regional probability filter (S + T)
  which(target_xy$specialist_check & target_xy$p_check),

  # land cover + regional probability filter (L + T)
  which(target_xy$lc_check & target_xy$p_check),

  # land cover + specialization + regional probability (S + L + T)
  which(target_xy$lc_check & target_xy$specialist_check & target_xy$p_check),

  # focus on specialist species (So)
  which(target_xy$n_habitats == 1),

  # combined filters (So + S + L + T)
  which(target_xy$lc_check & target_xy$specialist_check & 
          target_xy$p_check & (target_xy$n_habitats == 1))

)

# reduce observations to unique entries for each scenario ----
#-----------------------------------------------------------------------------#

# extract samples for each scenario
ols = list()
for (i in 1:length(ind)) {

  # source samples for scenario
  si = ind[[i]]
  if (length(si) == 0) next
  
  ols[[i]] = target_xy[si[!duplicated(target_xy$cell[si])],c("x","y")]
  ols[[i]]$group = groups[i]

  # estimate clustering weights (deals with spatial autocorr.; section 2.3.3)
  cell = cellFromXY(bioclim, as.matrix(ols[[i]][,c("x","y")]))
  ncell = ddply(data.frame(x=cell), .(x), summarize, n=length(x))
  wi = match(cell,ncell$x)
  ols[[i]]$weight = ncell$n[wi]
  ols[[i]]$weight = 1 / ols[[i]]$weight^2

}

# save samples
target_xy = do.call(rbind, ols)
rm(ols)

## extract and export bio-climatic data ----
##==========================================================================#

reference = extract(bioclim, reference_xy[,c("x","y")], ID=F)
reference = reference[complete.cases(reference),]
oname = file.path(o_path, paste0("0308_reference.csv"))
write.csv(reference, oname, row.names=F)

target = extract(bioclim, target_xy[,c("x","y")], ID=F)
ind = which(complete.cases(target))
target = target[ind,]

oname = file.path(o_path, paste0("0308_target.csv"))
write.csv(target, oname, row.names=F)

oname = file.path(o_path, paste0("0308_target_xy.csv"))
write.csv(target_xy, oname, row.names=F)

# number of subsamples to use
nss = nrow(reference)

```

### 2.3.1. Niche overlap analysis
```{r, eval=F}
# iterate through each scenario
gls = list()
pb = pb = txtProgressBar(min=1, max=length(groups), style=3)
for (i in 1:length(groups)) {

  # index of samples for scenario i
  ind = which(target_xy$group == groups[i])
  
  # skip if samples are insuficient
  # NOTE: # include to account for the small sample size of the example
  # (several 10s of thousands used in main manuscript)
  if (length(ind) < nss) next

  # define sub-sample size
  if (length(ind) > nrow(reference)) n = nrow(reference)

  # perform multiple runs to control for uncertainty
  p_ls = list()
  for (r in 1:niter) {

    # draw random sample
    ii = sample(ind, nss, prob=target_xy$weight[ind])

    # estimate niche overlap
    data = list(reference=niw.post(nsamples=ns, reference), 
                target=niw.post(nsamples=ns, target))
    no = overlap(data, nreps, nsamples, alpha=.95, norm.redraw=T)[2,1,]

    # preserve test results
    p_ls[[r]] = data.frame(group=groups[i],iteration=r, n=length(ind), p=no)

  }

  gls[[i]] = do.call(rbind, p_ls)
  setTxtProgressBar(pb, i)

}

close(pb)

# compile analyses
overlap = do.call(rbind, gls)

# return mean/sd for each scenario
stats = ddply(overlap, .(group), summarize, mean=mean(p), sd=sd(p))
```

### 2.3.2 Separability analysis
Note that the following code assumes that samples were generated for all the classes in the 'ecosystem' variable.
```{r}

# number of sub-sampling iterations
niter = 1000

# target ecosystems (this assumes more were processed)
ecosystems = c("0101","0104","0308","0105")

# load required data  ----
#-----------------------------------------------------------------------------#

# load metadata on samples
files = file.path(o_path, "samples", ecosystems, "_target_xy.csv")
obs = do.call(rbind, lapply(1:length(files), function(i) {
  tmp = read_csv(files[i], show_col_types=F)
  tmp$cid = i
  return(tmp)
}))

# load data required to apply the PCA model
files = paste0(o_path, "samples", ecosystems, "_target.csv")
pca = do.call(rbind, lapply(files, function(i) read_csv(i, show_col_types=F)))

# apply PCA model
pca = predict(pca_model, pca)

# analyse ecosystem separability ----
#-----------------------------------------------------------------------------#

separation = list()

for (r in 1:niter) {

  tmp = do.call(rbind, lapply(unique(obs$group), function(g) {

    ind = which(obs$group == g)

    si = unlist(lapply(unique(obs$cid), function(e) {
      i = ind[which(obs$cid[ind] == e)]
      sample(i, 100, prob=obs$weight[i])}))

    # calculate Davies-Bouldin's index
    o = index.DB(pca[si,], obs$cid[si], centrotypes="centroids", p=2, q=1)$DB

    # return data.frame
    return(data.frame(n=length(ind), db=o, group=g, iteration=r))

  }))

  separation[[r]] = tmp
  rm(tmp)
  print(r)

}

sds = do.call(rbind, separation)
write_csv(sds, file.path(o_path, "separability_analysis.csv"))

# return mean/sd for each scenario
stats = ddply(sds, .(group), summarize, mean=mean(db), sd=sd(db))
```
